---
title: "Lab 11 AlphaFold"
author: "Tiffany Chin 15700705"
format: html
---

## Structure Prediction using AlphaFold
Accurately predict protein 3D structure with its sequence using AlphaFold AI

Try AlphaFold using Adenylate Kinase

```{r}
library(bio3d)
id <- "1ake_A"
aa <- get.seq(id)
aa
```

```{r}
b <- blast.pdb(aa)
attributes(b)
```

```{r}
head(b$hit.tbl)
```

```{r}
hits <- plot(b)
```

```{r}
attributes(hits)
```

Top hits from blast results:

```{r}
hits$pdb.id
```

Now, download related PDB files

```{r}
files <- get.pdb(hits$pdb.id, path = "pdbs", split = TRUE, gzip = TRUE)
```

We have downloaded all ADK structures in the PDB database. To view efficiently, we need to align and superimpose these.

First, install the BiocManager package from CRAN. Using `BiocManager::install()`, we can install any bioconductor package. Then, we can align.

#install.packages("BiocManager")
#BiocManager::install("msa")

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Principal Component Analysis

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes= c(1:2))
```

Examine what PC2 (or any PCs) are capturing, we can plot the loadings to see the trajectory of the PCs

```{r}
mktrj(pc, pc=1, file="pc1.pdb")
```

## Class 11: Protein Structure Prediction with AlphaFold
6. Generating your own structure predictions the dimer input. As it is a homodimer this consists of the same sequence twice with a colon between chains (make sure to remove the spaces when you copy and paste into the line on the colab notebook):

>DB811864 full-length enriched swine cDNA library, adult ovary Sus scrofa cDNA clone OVRM10004A01 3', mRNA sequence
VESRDTPKDPAVTSKSPSMAQDSGPSELLPNGDLEKRSEPQPEEGSPAGGQKGGAPAEGE GAAETPSEASRAVENGCCTPKDGR

Follow #6-8 on the lab webpage: starting w/ 6. Visit AlphaFold2_mmseqs2 Colab note- book (at the time of writing this is currently the preferred AlphaFold version for our current prediction tasks.
